---
title: "Analysis Report 2: NEON Distribution of Elevation using Mosquito Trap Data"
author: "Alexia Ford"
date: "4/16/2018"
bibliography: references.bib
csl: ecology.csl
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This report is testing how the distribution of elevation varies across the National Ecological Observatory Network (NEON) sites and differing species, using data collected and publically presented by NEON. This report will explain NEON's work/purpose, NEON's design, test a hypothesis, and lastly analyze the results. The National Ecological Observatory Network is an observation platform for discovering, understanding, and forecasting the impacts of climate change, land use change, and invasive species on continental-scale ecology with sites distributed within 20 eco-climatic domains across United States, Alaska, Hawaii, and Puerto Rico using remote sensing technologies [@kampe2010neon]. NEON has allowed for easy accessible data to the public by providing the data the observatory has collected to scientists, land managers, policy makers, educators, and members of the general public interested in addressing these major ecological challenges. The data produced by NEON are strategically aimed at addressing ecological questions. Questions that are analyzing complex ecological processes and relationships that govern large scale patterns consistent with the measuring the drivers of change, biogeochemical cycling, and population and community responses across a variety of ecosystems in the ultimate hopes of expanding our understanding of ecological processes and to improve our ability to forecast these ecological changes [@thorpe2016introduction].

The project is responsible for measuring the causes and effects of environmental change across the United States within varying domains that represent different landforms, vegetation, climate, and ecosystem dynamics. The project is constantly collecting data to provide and statistically represent ecological, physical, and biological variability across the US. NEON has four different site types that the observatory collects data from, these sites include: core terrestrial sites, core aquatic sites, relocatable terrestrial sites, and lastly relocatable aquatic sites. The sites are collected  using five measurement subsystems of the Observatory including: the Airborne Observation Platform (AOP), Aquatic Instrument System (AIS), Aquatic Observation System (AOS), Terrestrial Instrument System (TIS), and Terrestrial Observation System (TOS) [@thorpe2016introduction]. These five systems of measurement have been used to collect endless data and NEON has continued to do so by properly standardizing protocols across all sites and maintaining community best practices. 

While NEON collects a wide range of data from atmospheric to even ecohydrology this report will specifically be looking at NEON's collection of mosquito data. The National Ecological Observatory Network (NEON) monitors mosquito populations across its broad geographical range of sites because of their prevalence in food webs, sensitivity to abiotic factors, and relevance for human health. Mosquitos (sentinel taxa) are helpful organisms to use in study because they are a diverse family of insects with varying life spans and generation times with wide geographic distributions that allow for comparisions across the continent as well as across major vegetation types. Mosquitos are of great importance to human health due to their epidemiological significance with associated diseases and their ecological influence of crops and healthy livestock. Mosquitos are also ideal for evaluating the ecological effects of global change as mosquitos are influenced by climate, vegetation, and host availability. Mosquitoes generally respond quickly to environmental change. Mosquitos are highly mobile and move to new areas depending on climatic conditions. For, example, as climate warms some mosquitos are expanding their geographical ranges and effeciency of transmission, resulting in mosquitos having more interactions with humans [@hoekman2016design]. 

NEON mosquito sampling takes place regularly during a time peroid in which mosquitos are active, this varies by domain. Sampling typically takes place more frequently at core sites than relocatable sites. Core sites are prioritized because of the long‐term nature of core site sampling. In addition, core sites are in wild land areas and provide baseline data to assess how ecological systems are changing through time [@hoekman2016design]. 

This report will specifically be analyzing how the distribution of elevation varies across all NEON sites and species using boxplots and geographical data. The elevation of the "ONAQ" NEON site and the elevation of *Culex tarsalis* and *Aedes nigromaculis* species will be of great significance for this report. This report will ask: what are some ways in which the distribution of elevation varies across NEON sites and species? Hypothesizing that the distribution of elevation of mosquitos at varying sites is dependent upon factors such as: feeding, vegetation, location, mating, etc. 

# Methods
The mosquito NEON data comes from a wide range of NEON sites with varying domains. The sampling design is continous across all sites to ensure data is consistent and comparable across time and space. The sampling methods include sampling mosquitos at NEON sites using CDC CO2 light traps. All traps use CO2 as bait because CO2 attracts the mosquitos as it is a component of vertebrate breath that mosquitos use to locate hosts. Once the mosquito is attracted to the trap a fan pulls the mosquitoes into a net bag for live storage until the trap is collected by a field technician. The design of the CDC CO2 light trap has remained intact since 1962, maintaining its effeciancy and effectiveness in use as the design has allowed for the removal of additional "trash" insects not swaying or adding to collection of data in anyway [@mcnelly1989cdc].

Once the trap is collected by a technician minimal processing occurs at the Domain lab. The mosquitos are frozen, transferred and stored in dry sample vials at -80°C to be sent for identification and pathogen testing at a facility. Mosquitos are identified at external facilities in order to minimize identification errors and accurately identiy mosquitos to species. Identification is based on visual examination of external morphology. After identification, target species are destructively tested for pathogens. 

Using mosquito data collected and provided by NEON this report will use R and RStudio to visualize results and test how the distribution of elevation varies across NEON sites and differing species. To begin, the mosquito NEON data was read into RStudio. R and RStudio was used in conjunction with the ggplot2 and dplyr packages to analyze the data. ANOVA was used in RStudio for plots made to test for the significance of the results found. The data, code, and other files needed to reproduce the analyses are available on GitHub. 

This report will specifically be analyzing how the distribution of elevation varies across all NEON sites and species using: the "ONAQ" NEON site, the elevation of *Culex tarsalis* species at all sites, and the elevation of *Aedes nigromaculis* species at all sites. The "ONAQ" NEON site was of significance for this report as it was at a higher elevation than other sites.

*Site Information*
"ONAQ" is short for Onaqui-Ault site. "ONAQ" is a core terrestrial site in Utah south of the rapidly developing Salt Lake City area. This area has been grazed by domestic livestock since settlement, the Forest Service has instituted a light to moderate fall and winter grazing regime. The vegetation of this site is dominantly sagebrush and western jupiner. Fire is a component of sagebrush, however invasions of exotic vegetation as well as spatial and temporal variability in resources has large‐scale effects such as the invasion of Bromus tectorum (aka drooping brome or cheatgrass) dramatically shortening fire return intervals [@chambers2007makes]. The invasion fire feedback has resulted in the loss of native perenial species allowing for the opportunity to specifically study the restoration of sagebrush at this site and collect data on the effects of invasive species and rodent dynamics. The climate of this region is dry as it is located in the intermountain region, between the Sierra Nevada and Cascade ranges. In winter, the climate is dominated by a cool moist air from the northern Pacific. The mountain ranges cast a rain shadow across the Intermountain Region resulting in low precipitation in the basins. Since this site is located in the Great Basin the site experiences the effects of dust from urban and agricultural use.  This domain is broadly characterized by combinations of lower-elevation, sagebrush steppe and higher-elevation montane forests. This complex topography impacts climate, hydrology, and species distributions, with montane zones exerting a disproportionate influence.

# Results
```{r load-packages}
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
```

```{r read-in-datasets}
# this is the information on each of the trapping locations and trapping events
mos_trapping <- read_csv("data/raw_data/mos_trapping.csv")

# this is the info about the taxonomic identity of mosquitos found in the traps
mos_expert <- read_csv("data/raw_data/mos_expertTaxonomistIDProcessed.csv")
```

```{r join-datasets}
# split the 'subsampleID' column into two new columns, including one called
# 'sampleID' that we can use to match to the other dataset
mos_expert <- mos_expert %>%
  separate(subsampleID,
           into = c("sampleID", "subsample_number"),
           sep = "\\.S\\.")

# join the two datasets together based on the shared 'sampleIS' column
mos_all <- mos_expert %>%
  full_join(mos_trapping, by = "sampleID", suffix = c(".taxa", ".sample"))
```

Figure 1. Box plot for the elevation of mosquitos at all NEON sites
```{r boxplot-of-mosquitos-elevation}
# elevation of mosquitos at all sites in the NEON data
ggplot(mos_all, aes(x = siteID.taxa,
                    y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Mosquitos at all NEON sites") +
  xlab("Site ID") +
  ylab("Elevation (meters)")
#r statistical test
anova(lm(elevation ~ siteID.taxa,
         data = mos_all))
```
**Chi-square test**
ANOVA p-value: 2.2e-16
This number shows that there are similarities in the elevation of mosquitos at all NEON sites because the number is so small there is a  very low precentage that what we are observing is random in this boxplot. For this reason we can reject the null hypothesis.

Figure 2. Box plot for the elevation of mosquitos at "ONAQ" NEON site
```{r boxplot-for-ONAQ-site}
# make a boxplot for the elevation of mosquitos at "ONAQ" NEON site
mos_all %>%
  filter(siteID.taxa == "ONAQ") %>%
  filter(!is.na(scientificName)) %>%
  ggplot(aes(x = scientificName,
             y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Mosquitos Species at ONAQ NEON site") +
  xlab("Species") +
  ylab("Elevation (meters)")
#r statistical test
anova(lm(elevation ~ scientificName,
         data = mos_all))
```
**Chi-square test**
ANOVA p-value: 2.2e-16
This number shows that there are similarities in the elevation of mosquito species at the "ONAQ" NEON site because the number is so small there is a  very low precentage that what we are observing is random in this boxplot. For this reason we can reject the null hypothesis.

Figure 3. Box plot for the elevation of *Culex tarsalis* at all NEON sites
```{r boxplot-of-culex-elevation-all-sites}
mos_all %>%
  filter(scientificName == "Culex tarsalis") %>%
  ggplot(aes(x = siteID.taxa,
             y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Culex tarsalis Species at all NEON Sites") +
  xlab("Site") +
  ylab("Elevation (meters)")
anova(lm(elevation ~ scientificName,
         data = mos_all))
```
**Chi-square test**
ANOVA p-value: 2.2e-16
This number shows that there are similarities in the elevation of *Culex tarsalis* species at all NEON sites because the number is so small there is a very low precentage that what we are observing is random in this boxplot. For this reason we can reject the null hypothesis.

Figure 4. Map of *Culex tarsalis* at all NEON sites
```{r map-of-culex-all-sites}
usa_map <- map_data("usa")

points_with_lat_long <- mos_all %>%
  filter(!is.na(decimalLatitude)) %>%
  filter(individualCount > 0) %>%
  filter(scientificName == "Culex tarsalis") %>%
  group_by(siteID.taxa) %>% 
  summarise(total_count = sum(individualCount),
            mean_lat = mean(decimalLatitude),
            mean_long = mean(decimalLongitude))

ggplot() +
  ggtitle("Elevation of Culex tarsalis Species at all NEON Sites") +
  geom_polygon(data = usa_map,
               aes(x = long,
                   y = lat,
                   group = group),
               fill = "light blue") +
  coord_fixed(1.3) +
  geom_point(data =  points_with_lat_long,
             aes(x = mean_long,
                 y = mean_lat,
                 size = total_count))
```

Figure 5. Box plot for the elevation of *Aedes nigromaculis* at all NEON sites
```{r boxplot-of-aedes-elevation-all-sites}
mos_all %>%
  filter(scientificName == "Aedes nigromaculis") %>%
  ggplot(aes(x = siteID.taxa,
             y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Aedes nigromaculis Species at all NEON Sites") +
  xlab("Site") +
  ylab("Elevation (meters)")
anova(lm(elevation ~ scientificName,
         data = mos_all))
```
**Chi-square test**
ANOVA p-value: 2.2e-16
This number shows that there are similarities in the elevation of *Aedes nigromaculis* species at all NEON sites because the number is so small there is a very low precentage that what we are observing is random in this boxplot. For this reason we can reject the null hypothesis.


Figure 1 is a boxplot correlating elevation and siteID for all NEON sites. Approximately 30 sites are shown in the figure with elevation showcased on a scale from 0 to 3,500 meters. Most of the sites elevation are not above 1,500 meters, with only three sites above 1,500 meters including: "CPER", "MOAB", and "ONAQ". Figure 2 models the elevation of varying species at the "ONAQ" site. Elevation is on a scale from 1600 to 1,950 meters. There are ten species showing elevation at the "ONAQ" site. All species present elevation: three species have lower elevations around 1,600 meteres, two species display higher elevations around 1,900 meters, and the other five species have elevations in the middle of the scale varying between 1,700 and 1,900 meters. Figure 3 displays the elevation of *Culex tarsalis* species at NEON sites. The elevation is showcased on a scale from 250 to 2,000 meters with a total of tweleve sites. The elevation for these twelve sites are varying with most either at high or low elevations and only one site at a middle elevation of 1,000 meters. Figure 4 is a United States map displaying a total count scale from 1,000 to 5,000 of *Culex tarsalis* species present at varying locations ploted in the US. The majority of *Culex tarsalis* present are seen in the middle of the map in the North and South regions, with no *Culex tarsalis* present in the East and some species present in the West. Figure 5 lastly, displays the elevation of *Aedes nigromaculis* species at NEON sites. The elevation is showcased on a scale from 250 to 1,750 meters with a total of nine sites. The elevation for these nine sites are varying with most either at high or low elevations. Four sites are shown above an elevation of 1,250 meteres while the other five sites are shown below and elevation of 750 meters. 

# Discussion
Figure 1 begins the report with a broad question: how does the distribution of elevation of mosquitos vary across all NEON sites? With this figure we see that the elevation is somewhat varying yet the majority of the elevation of mosquitos at approximately 30 sites are under 1,000 meters besides the sites: "CPER", "MOAB", and "ONAQ". Because of this another question was prosed, resulting in the creation of Figure 2: why is it that the "ONAQ" NEON Utah site is of that much greater elevation than most of the other NEON sites? This is further explained looking at a site in Colorado, the Northern Colorado Front Range, with a similiar elevation range of 1,500–2,400 meters. The study conducted shows that species richness was highest in plains habitats at elevations below 1,600 meters. Numerous species were found exclusively or predominantly at low elevations below 1,700 meters [@eisen2008mosquito]. Such as species found at the "ONAQ" NEON site, *Culex tarsalis* seen in Figure 3. Figure 3, 4, and 5 all prose a similar question: how does the elevation vary for mosquito species at the "ONAQ" site? Figure 3 we see that the elevation is typically either high (above 1,000 meters) or low (below 750 meters) with only one species seen between 750 and 1,000 meters. With this an interesting theme arose that landscape and elevation seem to be a pattern in the *Culex tarsalis* species. A study shows that host-seeking patterns in *Culex tarsalis* species indicated that passeriform birds (64%) and rabbits (25%) were the most frequent bloodmeal hosts and that the percentage of females feeding on birds varied temporally as an inverse function of mosquito abundance [@lothrop2001landscape]. Not only has it been studied that host-seeking patterns are prominent in *Culex tarsalis* species, its found that mating patterns are also temporally dependent as location and selevtion of an oviposition site is an essentional part of the life history of all mosquito species [@bentley1989chemical]. Figure 4, further showcases that sites with elevations between 1,000 meters to around 2,000 meters are more prominent in *Culex tarsalis* species abundance, as Northwest and Southwest regions of the United States are tyipcally in this elevation range, where as regions in the East have elevations closer to sealevel. Lastly, Figure 5, further answers the question: how does the elevation vary for mosquito species at the "ONAQ" site, by looking at the species *Aedes nigromaculis*. *Aedes nigromaculis* similarly to *Culex tarsalis* shows that the elevation for these nine sites *Aedes nigromaculis* is found at are varying with most either at high or low elevations. Interestingly so temporal location of these species matters. A study in the mountainous area of Wyoming show that the location of *Aedes nigromaculis* is dependent on both human abundance (lower elevation) and river location (high and low elevations). Locations populated with greater human abundance allows for the ample feeding of *Aedes nigromaculis* while on the other hand, being near a river is a preferable habit for the species [@wyoming1996elevational]. With this being said temporally both are benefically while areas in middle ground elevations are not as appealing to such species. INCLUDE VEGETATION OF ONAQ SITE IN RELATION TO ELEVATION. 

Overall, it seems that the hypothesis is in fact true in that the distribution of elevation of mosquitos at varying sites is dependent upon factors such as: feeding, vegetation, location, mating, etc. All these factors collectively result in the species abundance at NEON sites/elevations. 

# Sources Cited

