---
title: "Analysis Report 2: NEON Distribution of Elevation using Mosquito Trap Data"
author: "Alexia Ford"
date: "4/16/2018"
bibliography: references.bib
csl: ecology.csl
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This report is testing how the distribution of elevation varies across the National Ecological Observatory Network (NEON) sites and differing species, using data collected and publicly presented by NEON. This report will explain NEON's work/purpose, NEON's design, test a hypothesis, and lastly analyze the results. While NEON collects a wide range of data from atmospheric to even ecohydrology this report will specifically be looking at NEON's collection of mosquito data. The National Ecological Observatory Network (NEON) monitors mosquito populations across its broad geographical range of sites because of their prevalence in food webs, sensitivity to abiotic factors, and relevance for human health. Mosquitoes (sentinel taxa) are helpful organisms to use in study because they are a diverse family of insects with varying life spans and generation times with wide geographic distributions that allow for comparisons across the continent as well as across major vegetation types. Mosquitoes are of great importance to human health due to their epidemiological significance with associated diseases and their ecological influence of crops and healthy livestock. Mosquitoes are also ideal for evaluating the ecological effects of global change as mosquitoes are influenced by climate, vegetation, and host availability. Mosquitoes generally respond quickly to environmental change. Mosquitoes are highly mobile and move to new areas depending on climatic conditions. For, example, as climate warms some mosquitoes are expanding their geographical ranges and efficiency of transmission, resulting in mosquitoes having more interactions with humans [@hoekman2016design]. 

NEON mosquito sampling takes place regularly during a time period in which mosquitoes are active, this varies by domain. Sampling typically takes place more frequently at core sites than relocatable sites. Core sites are prioritized because of the long‐term nature of core site sampling. In addition, core sites are in wild land areas and provide baseline data to assess how ecological systems are changing through time [@hoekman2016design]. 

The National Ecological Observatory Network is an observation platform for discovering, understanding, and forecasting the impacts of climate change, land use change, and invasive species on continental-scale ecology with sites distributed within 20 eco-climatic domains across United States, Alaska, Hawaii, and Puerto Rico using remote sensing technologies [@kampe2010neon]. NEON has allowed for easy accessible data to the public by providing the data the observatory has collected to scientists, land managers, policy makers, educators, and members of the general public interested in addressing these major ecological challenges. The data produced by NEON are strategically aimed at addressing ecological questions. Questions that are analyzing complex ecological processes and relationships that govern large scale patterns consistent with the measuring the drivers of change, biogeochemical cycling, and population and community responses across a variety of ecosystems in the ultimate hopes of expanding our understanding of ecological processes and to improve our ability to forecast these ecological changes [@thorpe2016introduction].

Besides mosquito sampling NEON is responsible for measuring the causes and effects of environmental change across the United States within varying domains that represent different landforms, vegetation, climate, and ecosystem dynamics. The project is constantly collecting baseline data to provide and statistically represent ecological, physical, and biological variability across the US. NEON has four different site types that the observatory collects data from, these sites include: core terrestrial sites, core aquatic sites, relocatable terrestrial sites, and lastly relocatable aquatic sites. The sites are collected  using five measurement subsystems of the Observatory including: the Airborne Observation Platform (AOP), Aquatic Instrument System (AIS), Aquatic Observation System (AOS), Terrestrial Instrument System (TIS), and Terrestrial Observation System (TOS) [@thorpe2016introduction]. These five systems of measurement have been used to collect endless data and NEON has continued to do so by properly standardizing protocols across all sites and maintaining community best practices. 

This report will specifically be analyzing how the distribution of elevation varies across all NEON sites and species. The elevation of the "ONAQ" NEON site will be used as it was found to have the highest elevation and the elevation of *Culex tarsalis* and *Aedes nigromaculis* will be examined as these two species were found at the "ONAQ" NEON site. This report will ask: what are some ways in which the distribution of elevation varies across NEON sites and species? Hypothesizing that the distribution of elevation at varying sites is dependent upon factors such as: feeding, vegetation, location, mating, etc. and more specifically hypothesizing that high elevation sites would have a higher abundance of mosquitos.

# Methods

## Mosquito Sampling

The mosquito NEON data comes from a wide range of NEON sites with varying domains. The sampling design is continuous across all sites to ensure data is consistent and comparable across time and space. The sampling methods include sampling mosquitoes at NEON sites using CDC CO2 light traps. All traps use CO2 as bait because CO2 attracts mosquitoes as it is a component of vertebrate breath that mosquitoes use to locate hosts. Once the mosquito is attracted to the trap a fan pulls the mosquitoes into a net bag for live storage until the trap is collected by a field technician. The design of the CDC CO2 light trap has remained intact since 1962, maintaining it's efficiency and effectiveness in use as the design has allowed for the removal of additional "trash" insects not swaying or adding to collection of data in anyway [@mcnelly1989cdc].

Once the trap is collected by a technician minimal processing occurs at the Domain lab. The mosquitoes are frozen, transferred and stored in dry sample vials at -80°C to be sent for identification and pathogen testing at a facility. Mosquitoes are identified at external facilities in order to minimize identification errors and accurately identify mosquitoes to species. Identification is based on visual examination of external morphology. After identification, target species are destructively tested for pathogens. 

## Data Summarizing and Visualization

Using mosquito data collected and provided by NEON this report will use R and RStudio to visualize results and test how the distribution of elevation varies across NEON sites and differing species. To begin, the mosquito NEON data was read into RStudio. R and RStudio was used in conjunction with the ggplot2 and dplyr packages to analyze the data. 

## Statistical Analyses

ANOVA was used in RStudio for plots made to test for the significance of the results found. The data, code, and other files needed to reproduce the analyses are available on GitHub. 

## Site Information

I am focusing on one particular NEON site, ONAQ, because it is the highest in elevation. As my primary hypothesis is that high elevation sites will have higher abundance of mosquitos, an in-depth analysis of this site will allow me to test this hypothesis and research if there are in fact more mosquitos at this site due to high elevation. "ONAQ" is short for Onaqui-Ault site. "ONAQ" is a core terrestrial site in Utah south of the rapidly developing Salt Lake City area. This area has been grazed by domestic livestock since settlement, the Forest Service has instituted a light to moderate fall and winter grazing regime. The vegetation of this site is dominantly sagebrush and western jupiner. Fire is a component of sagebrush, however invasions of exotic vegetation as well as spatial and temporal variability in resources has large‐scale effects such as the invasion of Bromus tectorum (aka drooping brome or cheatgrass) dramatically shortening fire return intervals [@chambers2007makes]. The invasion fire feedback has resulted in the loss of native perennial species allowing for the opportunity to specifically study the restoration of sagebrush at this site and collect data on the effects of invasive species and rodent dynamics. The climate of this region is dry as it is located in the inter-mountain region, between the Sierra Nevada and Cascade ranges. In winter, the climate is dominated by a cool moist air from the northern Pacific. The mountain ranges cast a rain shadow across the Inter-mountain Region resulting in low precipitation in the basins. Since this site is located in the Great Basin the site experiences the effects of dust from urban and agricultural use.  This domain is broadly characterized by combinations of lower-elevation, sagebrush steppe and higher-elevation montane forests. This complex topography impacts climate, hydrology, and species distributions, with montane zones exerting a disproportionate influence.

# Results

```{r load-packages}
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
```

```{r read-in-datasets}
# this is the information on each of the trapping locations and trapping events
mos_trapping <- read_csv("data/raw_data/mos_trapping.csv")

# this is the info about the taxonomic identity of mosquitos found in the traps
mos_expert <- read_csv("data/raw_data/mos_expertTaxonomistIDProcessed.csv")
```

```{r join-datasets}
# split the 'subsampleID' column into two new columns, including one called
# 'sampleID' that we can use to match to the other dataset
mos_expert <- mos_expert %>%
  separate(subsampleID,
           into = c("sampleID", "subsample_number"),
           sep = "\\.S\\.")

# join the two datasets together based on the shared 'sampleIS' column
mos_all <- mos_expert %>%
  full_join(mos_trapping, by = "sampleID", suffix = c(".taxa", ".sample"))
```

Figure 1. Box plot for the elevation of mosquitoes at all NEON sites
```{r boxplot-of-mosquitos-elevation}
# elevation of mosquitos at all sites in the NEON data
ggplot(mos_all, aes(x = siteID.taxa,
                    y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Mosquitos at all NEON sites") +
  xlab("Site ID") +
  ylab("Elevation (meters)")
#r statistical test
anova(lm(elevation ~ siteID.taxa,
         data = mos_all))
```
**ANOVA Statistical Test**
ANOVA p-value: p < 0.001
This number shows that there are there are in fact significant differences between these NEON sites because the number is so small there is a very low percentage that what we are observing is random in this box plot. For this reason we can reject the null hypothesis.

## Figure 1 Description

I found that elevation significantly differed between NEON sites (Figure 1; ANOVA p < 0.001). Figure 1 is a box plot correlating elevation and siteID for all NEON sites. Approximately 30 sites are shown in the figure with elevation showcased on a scale from 0 to 3,500 meters. Most of the sites elevation are not above 1,500 meters, with only three sites above 1,500 meters including: "CPER", "MOAB", and "ONAQ". 

Figure 2. Box plot for the elevation of mosquitoes at "ONAQ" NEON site
```{r boxplot-for-ONAQ-site}
# make a boxplot for the elevation of mosquitos at "ONAQ" NEON site
mos_all %>%
  filter(siteID.taxa == "ONAQ") %>%
  filter(!is.na(scientificName)) %>%
  ggplot(aes(x = scientificName,
             y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Mosquitos Species at ONAQ NEON site") +
  xlab("Species") +
  ylab("Elevation (meters)")
#r statistical test
anova(lm(elevation ~ scientificName,
         data = mos_all))
```
**ANOVA Statistical Test**
ANOVA p-value: p < 0.001
This number shows that there are differences in the elevation of mosquito species at the "ONAQ" NEON site because the number is so small there is a  very low percentage that what we are observing is random in this box plot. For this reason we can reject the null hypothesis.

## Figure 2 Description

I found that elevation significantly differed between mosquito species at the "ONAQ" NEON site (Figure 2; ANOVA p < 0.001). Figure 2 models the elevation of varying species at the "ONAQ" site. Elevation is on a scale from 1,600 to 1,950 meters. There are ten species showing elevation at the "ONAQ" site. All species present elevation: three species have lower elevations around 1,600 meters, two species display higher elevations around 1,900 meters, and the other five species have elevations in the middle of the scale varying between 1,700 and 1,900 meters. 

Figure 3. Box plot for the elevation of *Culex tarsalis* at all NEON sites
```{r boxplot-of-culex-elevation-all-sites}
# make a boxplot for the culex species at all NEON sites using the
# mos_all data to make a ggplot and filtering out the culex species
mos_all %>%
  filter(scientificName == "Culex tarsalis") %>%
  ggplot(aes(x = siteID.taxa,
             y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Culex tarsalis Species at all NEON Sites") +
  xlab("Site") +
  ylab("Elevation (meters)")
anova(lm(elevation ~ scientificName,
         data = mos_all))
```
**ANOVA Statistical Test**
ANOVA p-value: p < 0.001
This number shows that there are differences in the elevation of *Culex tarsalis* species at all NEON sites because the number is so small there is a very low percentage that what we are observing is random in this box plot. For this reason we can reject the null hypothesis.

## Figure 3 Description

I found that elevation significantly differed between *Culex tarsalis* species at all NEON sites (Figure 3; ANOVA p < 0.001). Figure 3 displays the elevation of *Culex tarsalis* species at NEON sites. The elevation is showcased on a scale from 250 to 2,000 meters with a total of twelve sites. The elevation for these twelve sites are varying with most either at high or low elevations and only one site at a middle elevation of 1,000 meters. 

Figure 4. Map of *Culex tarsalis* at all NEON sites
```{r map-of-culex-all-sites}
# make a map using ggplot, the longitude and latitude values
# will be the x and y axis for the count of culex speices
usa_map <- map_data("usa")

points_with_lat_long <- mos_all %>%
  filter(!is.na(decimalLatitude)) %>%
  filter(individualCount > 0) %>%
  filter(scientificName == "Culex tarsalis") %>%
  group_by(siteID.taxa) %>%
  summarise(total_count = sum(individualCount),
            mean_lat = mean(decimalLatitude),
            mean_long = mean(decimalLongitude))

ggplot() +
  ggtitle("Elevation of Culex tarsalis Species at all NEON Sites") +
  geom_polygon(data = usa_map,
               aes(x = long,
                   y = lat,
                   group = group),
               fill = "light blue") +
  coord_fixed(1.3) +
  geom_point(data =  points_with_lat_long,
             aes(x = mean_long,
                 y = mean_lat,
                 size = total_count))
```
## Figure 4 Description
Figure 4 is a United States map displaying a total count scale from 1,000 to 5,000 of *Culex tarsalis* species present at varying locations plotted in the US. The majority of *Culex tarsalis* present are seen in the middle of the map in the North and South regions, with no *Culex tarsalis* present in the East and some species present in the West. 

Figure 5. Box plot for the elevation of *Aedes nigromaculis* at all NEON sites
```{r boxplot-of-aedes-elevation-all-sites}
# make a boxplot for the aedes species at all NEON sites using the
# mos_all data to make a ggplot and filtering out the aedes species
mos_all %>%
  filter(scientificName == "Aedes nigromaculis") %>%
  ggplot(aes(x = siteID.taxa,
             y = elevation)) +
  geom_boxplot() +
  ggtitle("Elevation of Aedes nigromaculis Species at all NEON Sites") +
  xlab("Site") +
  ylab("Elevation (meters)")
anova(lm(elevation ~ scientificName,
         data = mos_all))
```
**ANOVA Statistical Test**
ANOVA p-value: p < 0.001
This number shows that there are differences in the elevation of *Aedes nigromaculis* species at all NEON sites because the number is so small there is a very low percentage that what we are observing is random in this box plot. For this reason we can reject the null hypothesis.

## Figure 5 Description

I found that elevation significantly differed between *Aedes nigromaculis* species at all NEON sites (Figure 5; ANOVA p < 0.001). Figure 5 lastly, displays the elevation of *Aedes nigromaculis* species at NEON sites. The elevation is showcased on a scale from 250 to 1,750 meters with a total of nine sites. The elevation for these nine sites are varying with most either at high or low elevations. Four sites are shown above an elevation of 1,250 meters while the other five sites are shown below an elevation of 750 meters.

# Discussion

This report will specifically be analyzing how the distribution of elevation varies across all NEON sites and species. The "ONAQ" site will be of great importance to this report as well as the two species *Culex tarsalis* and *Aedes nigromaculis*. In the results section it was found that the elevation of the "ONAQ" NEON site had the highest elevation and the elevation of *Culex tarsalis* and *Aedes nigromaculis* species will be examined as these two species were found at the "ONAQ" NEON site. After analyzing the results of the distribution of elevation of NEON sites and distribution of species of mosquitoes at varying sites, further research will be conducted to see if other variables affect the distrubuition of elevation and distribution of species such as: feeding, vegetation, location, mating, etc. The primary hypothesized prosed is that high elevation sites would have a higher abundance of mosquitos.

Figure 1 begins the report with a broad question: how does the distribution of elevation vary across all NEON sites? With this figure we see that the elevation is somewhat varying yet the majority of the elevation of mosquitoes at approximately 30 sites are under 1,000 meters excluding the sites: "CPER", "MOAB", and "ONAQ". Because of this another question was posed, resulting in the creation of Figure 2: why is it that the "ONAQ", Onaqui-Ault, NEON Utah site is of that much greater elevation than most of the other NEON sites? This is further explained looking at a study done in Colorado, at the Northern Colorado Front Range, with a similar elevation range of 1,500–2,400 meters. The study conducted shows that species richness was highest in plains habitats at elevations below 1,600 meters. Numerous species were found exclusively or predominantly at low elevations below 1,700 meters [@eisen2008mosquito]. Such as the species found at the "ONAQ" NEON site, *Culex tarsalis* seen in Figure 3. Figure 3, 4, and 5 all prose a similar question: how does the elevation vary for mosquito species at the "ONAQ" site? In figure 3 we see that the elevation is typically either high (above 1,000 meters) or low (below 750 meters) with only one species seen between 750 and 1,000 meters. With this an interesting theme arose that landscape and elevation seem to be a pattern in the *Culex tarsalis* species. A study shows that host-seeking patterns in *Culex tarsalis* species indicated that passeriform birds (64%) and rabbits (25%) were the most frequent bloodmeal hosts and that the percentage of females feeding on birds varied temporally as an inverse function of mosquito abundance [@lothrop2001landscape]. Not only has it been studied that host-seeking patterns are prominent in *Culex tarsalis* species, its found that mating patterns are also temporally dependent as location and selection of an oviposition site is an essential part of the life history of all mosquito species [@bentley1989chemical]. Figure 4, further showcases that sites with elevations between 1,000 meters to around 2,000 meters are more prominent in *Culex tarsalis* species abundance, as Northwest and Southwest regions of the United States are typically in this elevation range, where as regions in the East have lower elevations, closer to sea level. Lastly, Figure 5, further answers the question: how does the elevation vary for mosquito species at the "ONAQ" site, by looking at the species *Aedes nigromaculis*. *Aedes nigromaculis* similarly to *Culex tarsalis* exhibit that the elevation for these nine sites with *Aedes nigromaculis* present are varying with most species abundance found at either high or low elevations. Interestingly so, temporal location of these species also matters. A study in the mountainous area of Wyoming show that the location of *Aedes nigromaculis* is dependent on both human abundance (lower elevation) and river location (high and low elevations). Locations populated with greater human abundance allows for the ample feeding of *Aedes nigromaculis* while on the other hand, being near a river is a preferable habit for the species [@wyoming1996elevational]. With this being said temporally both locations (high and low) are beneficial while areas in middle ground elevations are not as appealing to such species. To conclude, vegetation is also shown to have an effect on the sampling abundance of both *Culex tarsalis* and *Aedes nigromaculis* (both found at the "ONAQ" Utah NEON site).The study finds a relationship between vegetation and habitat of the *Culex tarsalis* and *Aedes nigromaculis* host-feeding patterns evaluated at the base of the Sierra Nevada foothills in Coachella Valley California, and also seen at the "ONAQ" NEON site [@meyer1991influence]. 

Overall, it seems that the hypothesis is in fact true in both results and study that high elevation sites would have a higher abundance mosquitos. Resarch has also allowed for the conclusion that the distribution of elevation across varying sites is dependent upon factors such as: feeding, vegetation, location, mating, etc. All these factors collectively result in the varying species abundance at NEON sites/elevations. 

# Sources Cited

